---
title: "Health Expectancy"
author: |
  | Tim Riffe
  | MPIDR
date: "18 Nov, 2021"
output: 
  beamer_presentation:
    colortheme: "frigatebird"
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{arrows,calc,positioning,shapes.geometric}
  - \usepackage{pgfplots}
  - \pgfplotsset{compat=1.15}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## 

\Large
> Average number of years lived in good health

\pause

> Summary index of an underlying distribution

\pause

> What's *health* anyway?

\pause

> model as a stock or a process?

<!-- ## Motivation -->

<!-- - Quality vs quantity of years? -->
<!-- - Hidden inequality? -->
<!-- - Process? -->
<!-- - Labor force (GALI question)? -->
<!-- - Advocacy? -->

<!-- ## Monitoring -->

<!-- \begin{center} -->
<!-- \includegraphics[width=11cm]{Healthy-life-years-at-birth_04_04_20.png} -->
<!-- \end{center} -->

<!-- ## ... -->

<!-- \Large -->
<!-- > What about intervention? -->

<!-- ## Intervention? -->

<!-- Campell's Law -->

<!-- > “The more any quantitative social indicator is used for social decision-making, the more subject it will be to corruption pressures and the more apt it will be to distort and corrupt the social processes it is intended to monitor” -->

<!-- \footnotesize -->
<!-- Campbell, D. T. (1979). Assessing the impact of planned social change. Evaluation and program planning, 2(1), 67-90. -->

<!-- \pause -->
<!-- Goodhart's Law -->

<!-- > “Any observed statistical regularity will tend to collapse once pressure is placed upon it for control purposes.” -->

<!-- \footnotesize -->
<!-- Goodhart, Charles (1975). "Problems of Monetary Management: The U.K. Experience". Papers in Monetary Economics. 1. Sydney: Reserve Bank of Australia. -->


## Measures

\Huge
- Sullivan
- Multistate

## Sullivan 

```{r, message = FALSE, warn = FALSE, echo = FALSE, out.width='70%', out.height='70%'}
library(tidyverse)
library(ggdark)
pix <- 1-c(0.942832933525023, 0.925366777629796, 0.913878052953308, 0.905779576715131, 
         0.899819586082213, 0.895228598060479, 0.891415053920396, 0.887843431788832, 
         0.883977734168754, 0.879243670749217, 0.872988101030959, 0.864422353013589, 
         0.85253805474243, 0.836000848599391, 0.813416659727759, 0.783638907018477, 
         0.745769572698891, 0.699353730294645, 0.64468449536391, 0.583095062238541, 
         0.5170619612345, 0.449955803009552, 0.385423852181739, 0.326631969263549, 
         0.275728924251826, 0.23377510423577, 0.201092734009695, 0.17782145726523, 
         0.164539328404487, 0.163055237515445)

lx <- c(2, 1.98744304104249, 1.97166480704484, 1.95251099982402, 1.92979047491913, 
  1.90327639864472, 1.87269654546685, 1.83771283501313, 1.7978930617532, 
  1.75267423877303, 1.70131782289022, 1.64285829147167, 1.57605111088635, 
  1.49933944580084, 1.41094952851575, 1.30916546330524, 1.19275620234035, 
  1.06156436022103, 0.917173279730504, 0.763469971998791, 0.606832911763207, 
  0.455676937888572, 0.319219834302956, 0.205583276597247, 0.119696424658899, 
  0.0618521812086196, 0.0278148278482655, 0.0106670648655719, 0.00341876365030296, 
  0.000898335418648003) / 2

tibble(age = seq(50,108,by=2),
       pix = pix,
       lx = lx) %>% 
  ggplot(aes(x= age, y = lx)) +
  geom_line(color = "white") +
  geom_line(inherit.aes = FALSE, mapping = aes(x=age,y = pix), color = "red") +
 dark_theme_gray() + 
  theme(
        plot.background = element_rect(fill = "grey10"),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey30", size = 0.2),
        panel.grid.minor = element_line(color = "grey30", size = 0.2),
        legend.background = element_blank(),
        axis.ticks = element_blank(),
        legend.key = element_blank(),
        axis.title = element_text(size=14,face="bold"),
        axis.text =  element_text(size = 12))
# a <- seq(50,108,by=2)
# # survival curve
# pdf("test.pdf",width = 5, height = 5)
# par(mai=c(.5,.6,.1,.1))
# plot(a, lx, type = 'n', las = 2,
#      bg="white",
#      col.main="white",
#      xlab = "",
#      axes = FALSE,
#      ylab = "")
# axis(1,col = "white",col.ticks = "white",col.axis ="white")
# axis(2,col = "white",col.ticks="white",col.axis ="white",las=1)
# # color part of it healthy
# lines(a, 1- pix, col = "red", lwd = 2)
# lines(a, lx, col = "white", lwd = 2)
# dev.off()
```

<!-- \begin{center} -->
<!-- \includegraphics[scale=.7]{test.pdf} -->
<!-- \end{center} -->

- $HLE(x) = \int l(x)\pi(x) \mathrm{d}x$
- List properties \& biases (5 min)

## Multistate

<https://temery86.github.io/FullHistory/>

\begin{center}
\includegraphics[width=11cm]{multistate_animation.png}
\end{center}

## Multistate

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse)
library(ggdark)
read_csv(here::here("Data","TRv6.csv")) %>% 
filter(sex == "f",
       edu == "terciary",
       time == 1996) %>% 
pivot_longer(m11:m24, 
             names_to = "transition", 
             values_to = "probability") %>% 
mutate(transition = case_when(
  transition == "m11" ~ "H->H",
  transition == "m12" ~ "H->U",
  transition == "m14" ~ "H->D",
  transition == "m21" ~ "U->H",
  transition == "m22" ~ "U->U",
  transition == "m24" ~ "U->D",
)) %>% 
  separate(transition, into = c("from","to"),sep = "->") %>% 
  ggplot(aes(x = age,
             y = probability,
             color = to, 
             lty = from,
             group = interaction(from,to))) +
  geom_line(size = 2) +
  dark_theme_gray() + 
  theme(
        plot.background = element_rect(fill = "grey10"),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey30", size = 0.2),
        panel.grid.minor = element_line(color = "grey30", size = 0.2),
        legend.background = element_blank(),
        axis.ticks = element_blank(),
        legend.key = element_blank(),
        axis.title = element_text(size=14,face="bold"),
        axis.text =  element_text(size = 12))

```

## State space-diagrams

\small
\begin{center}
\begin{tikzpicture}[scale=0.45, ->,>=stealth',shorten >=1pt,auto,node distance=2.5cm,
  thick,main node/.style={circle,draw},square/.style={regular polygon,regular polygon sides=4}]

  \node[main node] (1) at (0,0) [square,draw]{Healthy};
  \node[main node] (2) at (8,-5) [square,draw] {Disabled};
  \node[main node] (3) at (3,-12) [square,draw] {Dead};  

  \path[every node/.style={font=\sffamily\small}]
    
    (1) edge node [left] {} (3)
   	(1) edge [bend right] node [left] {} (2)
    
    (2) edge node [right] {} (3)
    (2) edge [bend right] node [left] {} (1)

    (1) edge [loop left] node {} (1)
    (2) edge [loop right] node {} (2)
    (3) edge [loop below] node {} (3)
 
    ;
\end{tikzpicture}
\end{center}


## Properties

\begin{center}
\includegraphics[width=11cm]{chuttersnap-gts_Eh4g1lk-unsplash.jpg}
\end{center}

<!-- ## Properties -->

<!-- \begin{center} -->
<!-- \includegraphics[width=11cm]{josh-withers-3mY4U7FIevc-unsplash.jpg} -->
<!-- \end{center} -->

<!-- ## Properties -->

<!-- \begin{center} -->
<!-- \includegraphics[width=11cm]{nii-sx7kn7GNf6E-unsplash.jpg} -->
<!-- \end{center} -->

## Variants

- Discrete
  - Matrix algebra *
  - Lifetable *
  - Simulation *
- Continuous
  - Matrix algebra
  - Differential equations
  
## Considerations

- Initial conditions?
- Mid-interval approximations?
- Strata or state?
- Memoryless?
- Rate fluctuations?
- Health comparability?
  
## Equivalencies

- Cohorts
- Stationary populations
  
## Further training

- Jutta Gampe Multistate course possible late May (check PHDS website in March)
- Hal Caswell matrix pop course will eventually return
- REVES meetings have workshops

## Special conferences

- IAPHS <https://iaphs.org/>
- REVES <https://reves.site.ined.fr/en/home/about_reves/>

## Local technical resources

- Transition probs: J. Gampe, A. Carollo, D. Schneider, A. Lorenti, C. Dudel
- Markov measures: C. Dudel, me, C. Bohk-Ewald

## Related things that I work on

- Within-inequality in Sullivan setting
- Decomposition of discrete multistate
- Other within-lifecourse duration measures
- Time perspectives on compression and ageing

## Break, then we code!

\begin{center}
\includegraphics[width=9cm]{is_it_worth_the_time.png}
\end{center}
